---
title: "smart city characteristics"
author: "cm"
date: "2024-01-11"
output: html_document
---


```{r}
options(scipen=16)
#library(rstatix)
library(tidyverse)
```

```{r}
df <- read_csv("data/smart_cities20240110.csv")
ds254 <- read_csv("data/nhgis/nhgis0017_csv/nhgis0017_ds254_20215_place.csv")
ds255 <- read_csv("data/nhgis/nhgis0017_csv/nhgis0017_ds255_20215_place.csv")
ds248 <- read_csv("data/nhgis/nhgis0018_csv/nhgis0018_ds248_2020_place.csv")
ds254_2 <- read_csv("data/nhgis/nhgis0018_csv/nhgis0018_ds254_20215_place.csv")
ds254_3 <- read_csv("data/nhgis/nhgis0019_csv/nhgis0019_ds254_20215_place.csv")
```

```{r}
df <- df %>% mutate(pop_size=case_when(POP20>=1000000 ~ 4, POP20<1000000&POP20>=100000 ~3, POP20<100000&POP20>= 10000 ~2, POP20<10000 ~1)) %>% mutate(pop_label=case_when(POP20>=1000000 ~ "Population: 1 million or more", POP20<1000000&POP20>=100000 ~ "Population: 100,000 to 999,999", POP20<100000&POP20>= 10000 ~"Population: 10,000 to 99,999", POP20<10000 ~"Population: less than 10,000")) 
df$pop_label <- factor(df$pop_label, levels = c("Population: 1 million or more", "Population: 100,000 to 999,999", "Population: 10,000 to 99,999", "Population: less than 10,000"))
```

```{r}
mycolor <- c("#596174","#ea6a47" )
```

# cleaning the data

```{r}
ds254 <- ds254 %>% 
# Percentage of commuters 16 and over carpooling, using public transport, bicycles, or walking to work
# Data: Table 1:     Means of Transportation to Work
  mutate(green_commute = (AOOTE004+AOOTE010+AOOTE018+AOOTE019)/(AOOTE001-AOOTE021)*100, 
# Percentage of the population 25 and over with a bachelor’s degree or higher
# Data: Table 2:     Educational Attainment for the Population 25 Years and Over
        bachelor = (AOP8E022+AOP8E023+AOP8E024+AOP8E025)/AOP8E001*100, 
# Percentage of households living below poverty line
# Data: Table 3:     Poverty Status in the Past 12 Months by Household Type by Age of Householder
        poverty = AOQEE002/AOQEE001*100,
# Per capita income
# Data: Table 4:     Per Capita Income in the Past 12 Months (in 2021 Inflation-Adjusted Dollars)
        income = AORME001,
# Median value of housing units
# Data: Table 7:     Median Value (Dollars)
        housing_value = AOULE001,
# Percentage of the population without health insurance
# Data: Table 9:     Types of Health Insurance Coverage by Age
        no_insurance = (AOU0E017+AOU0E033+AOU0E050+AOU0E066)/AOU0E001*100, 
# Percentage of households with internet subscription
# Data: Table 10:    Presence and Types of Internet Subscriptions in Household
        internet = AOU2E002/AOU2E001*100,
# Percentage of households with computers
# Data: Table 11:    Computers in Household
        computer = (AOVIE001-AOVIE007)/AOVIE001*100) %>% 
  select(GISJOIN, green_commute, bachelor, poverty, income, housing_value, no_insurance, internet, computer)
```

```{r}
ds255 <- ds255 %>% 
# Gini index of income inequality (It measures the degree of income inequality on a 0-1 scale. The more equal the income distribution, the lower the Gini index
# Data: Table 1:     Gini Index of Income Inequality
  mutate(gini = AO72E001) %>% select(GISJOIN, gini)
```

```{r}
ds248 <- ds248 %>% 
# Percentage of vacant housing units
# Data: Table 1:     Occupancy Status
  mutate(vacant_rate = U7G003/U7G001*100) %>% select(GISJOIN,vacant_rate)
```

```{r}
ds254_2 <- ds254_2 %>% 
# Median age
# Data: Table 1:     Median Age by Sex
  mutate(age = AONUE001) %>% select(GISJOIN, age)
```

```{r}
ds254_3 <- ds254_3 %>% 
# Racial and ethnic diversity index: 1-(H2+NHW2+NHB2+NHA2+NHNH2+NHO2+NHTOM2)
# https://www.census.gov/newsroom/blogs/random-samplings/2021/08/measuring-racial-ethnic-diversity-2020-census.html
# - Hispanic
# - White alone, non-Hispanic
# - Black or African American alone, non-Hispanic
# - American Indian and Alaska Native alone, non-Hispanic
# - Asian alone, non-Hispanic
# - Native Hawaiian and Other Pacific Islander alone, non-Hispanic
# - Some Other Race alone, non-Hispanic
# - Multiracial, non-Hispanic
# "This index shows the probability that two people chosen at random will be from different race and ethnic groups."
# "The DI is bounded between 0 and 1, with a zero-value indicating that everyone in the population has the same racial and ethnic characteristics, while a value close to 1 indicates that everyone in the population has different characteristics."
# Data: Table 2:     Hispanic or Latino Origin by Race
  mutate(diversity = 1-((AOOCE012/AOOCE001)^2+(AOOCE003/AOOCE001)^2+(AOOCE004/AOOCE001)^2+(AOOCE005/AOOCE001)^2+(AOOCE006/AOOCE001)^2+(AOOCE007/AOOCE001)^2+(AOOCE008/AOOCE001)^2+(AOOCE009/AOOCE001)^2),
# Unemployed rate: (Unemployed/Labor Force) x 100
# Data: Table 4:     Employment Status for the Population 16 Years and Over
        unemp = AOSJE005/AOSJE002*100,
# Percentage male
# Data: Table 1:     Sex by Age
        male = AONTE002/AONTE001*100) %>% 
  select(GISJOIN, diversity, unemp, male)
```

```{r}
df <- merge(df, ds254, by = "GISJOIN", all.x = TRUE)
df <- merge(df, ds255, by = "GISJOIN", all.x = TRUE)
df <- merge(df, ds248, by = "GISJOIN", all.x = TRUE)
df <- merge(df, ds254_2, by = "GISJOIN", all.x = TRUE)
df <- merge(df, ds254_3, by = "GISJOIN", all.x = TRUE)
```

# plot functions

```{r}
plot_mean <- function(data, x, y, lab_text) {
  data_mean <- data %>% 
    group_by({{x}}) %>% 
    summarise(n=n(), mean = round(mean({{y}}, na.rm=T),2), .groups = "drop")
  print(data_mean)
  
  data_plot <- data %>% drop_na({{y}})

  ggplot() +
    geom_density(data_plot, mapping=aes({{y}}, fill=as.factor({{x}})), alpha=0.4, color = NA) +
    geom_vline(data_mean, mapping=aes(xintercept=mean, color = as.factor({{x}})),linetype="dashed") +
    #geom_text(data_mean, mapping=aes(mean,+Inf, label=round(mean,1), color=as.factor(smart)),angle=90, vjust=1, hjust=2, size=3) +
    scale_fill_manual(name=" ", labels=c("1"="Smart cities", "0"="Non-smart cities"), values = mycolor) +
    scale_color_manual(values = mycolor, guide="none") +
    theme_classic() +
    xlab(lab_text) +
    ylab("Density") 
}
# The height of the curve is density, not probability. Probability is represented by area: multiplying density by an infinitesimal width along the x-axis to get the probability of variable x.
```

```{r}
plot_facet <- function(data, x, y, lab_text) {
  data_mean <- data %>% 
    group_by( pop_label, {{x}}) %>% 
    summarise(n=n(), mean = round(mean({{y}}, na.rm=T),2), .groups = "drop")
  print(data_mean)
  
  data_plot <- data %>% drop_na({{y}})

  ggplot() +
    geom_density(data_plot, mapping=aes({{y}}, fill=as.factor({{x}})), alpha=0.4, color = NA) +
    geom_vline(data_mean, mapping=aes(xintercept = mean, color = as.factor({{x}})),linetype="dashed") +
    scale_fill_manual(name=" ", labels=c("1"="Smart cities", "0"="Non-smart cities"), values = mycolor) +
    scale_color_manual(values = mycolor, guide="none") +
    theme_bw()+
    xlab(lab_text) +
    ylab("Density") +
    facet_wrap(~ pop_label)
}
```

## Percentage male
```{r}
plot_mean(df, smart, male, "Percentage male")
```

```{r}
plot_facet(df, smart, male, "Percentage male")
```
## Median age

```{r}
plot_mean(df, smart, age, "Median age")
```

```{r}
plot_facet(df, smart, age, "Median age")
```
## !Racial and ethnic diversity index

```{r}
plot_mean(df, smart, diversity, "Racial and ethnic diversity index")
```

```{r}
plot_facet(df, smart, diversity, "Racial and ethnic diversity index")
```

## !Percentage of the population 25 and over with a bachelor's degree or higher


```{r}
plot_mean(df, smart, bachelor, "Percentage of population 25 and over with a bachelor’s degree or higher")
```

```{r}
plot_facet(df, smart, bachelor, "Percentage of population 25 and over with a bachelor’s degree or higher")
#ggsave("figure/edu.png")
```

## !Per capita income

```{r}
plot_mean(df, smart, income, "Per capita income (dollar)")
```
```{r}
41029.94-31005.14
```

```{r}
plot_facet(df, smart, income, "Per capita income (US dollar)")
```
```{r}
37556.12-34520.57
42698.93-36487.84
46138.68-30089.62
```

```{r}
plot_facet2 <- function(data, x, y, lab_text) {
  data_mean <- data %>% 
    group_by( pop_label, {{x}}) %>% 
    summarise(n=n(), mean = round(mean({{y}}, na.rm=T),2), .groups = "drop")
  print(data_mean)
  
  data_plot <- data %>% drop_na({{y}})

  ggplot() +
    geom_density(data_plot, mapping=aes({{y}}, fill=as.factor({{x}})), alpha=0.4, color = NA) +
    geom_vline(data_mean, mapping=aes(xintercept = mean, color = as.factor({{x}})),linetype="dashed") +
    scale_fill_manual(name=" ", labels=c("1"="Smart cities", "0"="Non-smart cities"), values = mycolor) +
    scale_color_manual(values = mycolor, guide="none") +
    scale_x_continuous(limits = c(0, 100000)) +
    theme_bw()+
    xlab(lab_text) +
    ylab("Density") +
    facet_wrap(~ pop_label)
}
```
```{r}
plot_facet2(df, smart, income, "Per capita income (US dollar)")
```
Note: The plot does not display per capita income exceeding 100,000 US dollars.


## !Gini index of income inequality

```{r}
plot_mean(df, smart, gini, "Gini index of income inequality")
```

```{r}
plot_facet(df, smart, gini, "Gini index of income inequality")
#ggsave("figure/gini.png")
```

## !Median value of housing units

```{r}
plot_mean(df, smart, housing_value, "Median value of housing units")
```

```{r}
plot_facet(df, smart, housing_value, "Median value of housing units")
```

## Percentage of vacant housing units

```{r}
plot_mean(df, smart, vacant_rate, "Percentage of vacant housing units")
```

```{r}
plot_facet(df, smart, vacant_rate, "Percentage of vacant housing units")
```

## Percentage of commuters 16 and over carpooling, using public transport, bicycles, or walking to work
```{r}
plot_mean(df, smart, green_commute, "Percentage of commuters carpooling, using public transport, bicycles, or walking to work")
```

```{r}
plot_facet(df, smart, green_commute, "Percentage of commuters carpooling, using public transport, bicycles, or walking to work")
#ggsave("figure/transportation.png")
```
## Unemployed rate

```{r}
plot_mean(df, smart, unemp, "Percentage of population umemployed")
```

```{r}
plot_facet(df, smart, unemp, "Percentage of population umemployed")
#ggsave("figure/employment.png")
```

## Percentage of population without health insurance

```{r}
plot_mean(df, smart, no_insurance, "Percentage of population without health insurance")
```

```{r}
plot_facet(df, smart, no_insurance, "Percentage of population without health insurance")
#ggsave("figure/insurance.png")
```

## Percentage of households living below poverty line

```{r}
plot_mean(df, smart, poverty, "Percentage of households living below poverty line")
```

```{r}
plot_facet(df, smart, poverty, "Percentage of households living below poverty line")
#ggsave("figure/poverty.png")
```

## !Percentage of households with internet subscription

```{r}
plot_mean(df, smart, internet, "Percentage of households with internet subscription")
```

```{r}
plot_facet(df, smart, internet, "Percentage of households with internet subscription")
#ggsave("figure/internet.png")
```

## !Percentage of households with computers

```{r}
plot_mean(df, smart, computer, "Percentage of households with computers")
```

```{r}
plot_facet(df, smart, computer, "Percentage of households with computers")
```

```{r}
write.csv(df, "data/acs20240404.csv", row.names=FALSE)
```
